Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-11-01 paper/paper.tex
Clean up comments
Minor
Pass through introduction

%%
%% This is file `sample-acmsmall.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[\ifpublic nolinenum\else\fi,online,OA]{jfp}

\let\Bbbk\relax
\usepackage{ottalt}
\usepackage[outputdir=latex.out]{minted}
\usepackage{xspace}
\usepackage{stmaryrd}
\usepackage{bbding}
\usepackage{amsthm,amssymb}
\usepackage{dejavu}
\usepackage[para]{footmisc}
\newcommand{\dotv}[2]{\href{#1}{\texttt{#1}}{\texttt{:#2}}}
\newcommand{\lang}{$\lambda^{\Pi}$\xspace}
\inputott{rules}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

\let\draftnote\relax % this is already defined in JFP
\usepackage{draft}

\ifcomments
\newnote{scw}{blue} % Stephanie Weirich
\newnote{yl}{purple} % Yiyun Liu
\else
\newcommand{\scw}[1]{}
\newcommand{\yl}[1]{}
\fi

@@

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% For managing citations, it is recommended to use bibliography
%% files in BibTeX format.
%%
%% You can then either use BibTeX with the ACM-Reference-Format style,
%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
%% support for advanced citation of software artefact from the
%% biblatex-software package, also separately available on CTAN.
%%
%% Look at the sample-*-biblatex.tex files for templates showcasing
%% the biblatex styles.
%%

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
\citestyle{acmauthoryear}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

\journaltitle{JFP}
\cpr{The Author(s),}
\doival{10.1017/}
\totalpg{\pageref{lastpage}}
\jnlDoiYr{2025}
\lefttitle{Y. Liu et al.}
\righttitle{Logical Relations for Type Theory}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\begin{authgrp}
\author{Yiyun Liu}
% \orcid{0009-0006-8717-2498}
\affiliation{University of Pennsylvania (\email{\elink{liuyiyun@seas.upenn.edu}})}

\author{Stephanie Weirich}
% \orcid{0000-0002-6756-9168}
\affiliation{University of Pennsylvania (\email{\elink{sweirich@seas.upenn.edu}})}

\author{Jonathan Chan}
\affiliation{University of Pennsylvania (\email{\elink{jcxz@seas.upenn.edu}})}
\end{authgrp}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
% \renewcommand{\shortauthors}{Trovato et al.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Proof by logical relations is a powerful technique that has been used
to derive metatheoretic properties of type systems, such as
consistency and parametricity. While there exists a
plethora of introductory materials about logical relations in the
context of simply typed or polymorphic lambda calculi, a streamlined
presentation of proof by logical relation for a dependently typed language
is lacking. In this paper, we present a short
consistency proof for a dependently typed language that contains a
rich set of features, including a countable universe
hierarchy, booleans, and an intensional identity type. We show that
the logical relation can be easily extended to prove the existence of
$\beta\eta$-normal forms.
We have
fully mechanized the consistency proof using the Coq proof assistant
in under 1000 lines of code, with 500 lines of additional code for the
$\beta\eta$-normal form extension.
\end{abstract}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle[T]

\section{Introduction}
This paper presents a \emph{short} and \emph{mechanized} proof of
logical consistency for \lang{}, a dependent type theory with a full
predicative universe hierarchy, large eliminations, an intensional identity
type, a boolean base type, and dependent elimination forms.

@@

However, tutorial material on syntactic logical
relations~\citep{skorstengaard2019introduction, harpertait, harperkripke,
  pierce2002types, pierce2004advanced,harper2016practical} is primarily
focused on systems with simple or polymorphic types. In that context,
syntactic logical relations can be defined as simple recursive functions over
the structure of types, or (in the case of recursive types) defined over the
evaluation steps of the computation. Yet, neither of these techniques can be
used to define a logical relation in the context of a predicative dependent
type theory, so a novice researcher might be excused for thinking that proofs
that use logical relations are not applicable for such languages.

But this is not the case. 
% \scw{Add a sentence about pen-and-paper LR proofs
%  for dependent type theories?}\yl{There is a long paragraph in Sec 8 about how
%  pnp proofs like Geuvers' can't be easily mechanized, but I'm not sure if we
%  can also argue that they are hard to read, too (I personally think that's the
%  case, but I don't know if the reviewers will agree)} 
Recent authors have developed tour-de-force
mechanizations for the metatheory of modern proof
assistants~\citep{nbeincoq,decagda,martin-lof-a-la-coq,anand2014towards}, and
have relied on logical relations defined as part of their
developments. However, because these proofs show diverse results about real
systems and algorithms, these developments range in size from 10,000 to
300,000 lines of code. As a result, their uses of logical relations are
difficult to isolate from the surrounding contexts and inaccessible
to casual readers.

Thus, our paper provides a gentle and accessible introduction to a powerful
technique for dependent type theories. To promote the use of machine-assisted
reasoning, our development is accompanied by a short mechanized proof script,
of less than 1000 lines of code, developed using the Coq proof
assistant~\cite{coq}.

We have streamlined our proof through a number of means: the careful selection
of the features that we include in the object type system and the results that
we prove about it, in addition to the judicious use of automation.  Our
language is small, but includes enough to be illustrative. For example, we
eschew inductive datatypes or W-types, but we do include propositional
equality and booleans to capture the challenges presented by indexed types and
dependent pattern matching. We do not show the decidability of type checking,
nor do we develop a PER semantics, but we prove logical consistency, which
states that empty types are not inhabited in an empty context, and demonstrate
how our consistency proof can be extended (at a moderate cost of 500 lines of
code) to show the existence of $\beta\eta$-normal forms for well-typed open
\emph{and} closed terms. We include a full predicative universe hierarchy and
type-level computation to demonstrate the logical strength of the approach.

@@

More concretely, our paper makes the following contributions.
\begin{itemize}
\item In Section~\ref{sec:spec}, we introduce \lang{}, the dependent type
  theory of interest. A key design choice that impacts our proofs is the use
  of an untyped conversion rule, inspired by Pure Type
  Systems~\citep{barendregt1991introduction}, and specified through parallel
  reduction~\citep{takahashi-parallel-reduction,
    barendregt:lambda-calculi-with-types}.
\item In Section~\ref{sec:logreldep}, we formulate logical consistency for
  \lang{}, the property of interest, to motivate a logical relation. We define
  the relation first inductively and then later prove that it is a partial
  function. Based on this definition, we define semantic typing and prove the
  fundamental theorem, from which consistency follows as a corollary
  (Section~\ref{sec:logrelproof}). Thanks to the design of \lang{}, our proof
  showcases the special treatment required to model many of the most common
  features of dependent type theories, thus making our proof applicable to a
  broad range of type systems.
\item We strengthen our logical relation to prove the existence of
  $\beta$-normal forms (Section~\ref{sec:extension}) and $\beta\eta$-normal
  forms (Section~\ref{sec:betaeta}) for well-typed open terms. The
  modifications made to our initial logical relation are small and closely
  mirror the necessary extensions for a simply-typed language. We use this
  part to show that once we have established the base techniques, we can port
  ideas from proofs about simpler languages to the dependently typed setting.
\item We mechanize all our proofs using the Coq proof assistant, with 957
  lines of code for the consistency proof and a moderate increase to 1568
  lines of code for the normalization proof.  We discuss the specifics related
  to our choice of Coq as our metatheory, including our use of off-the-shelf
  semantic engineering infrastructure and automation tools, in
  Section~\ref{sec:logrelmech}.  Our proof scripts, with comments, are
  available to reviewers as supplementary materials.
\item We compare our work to existing proofs by logical relations and other
  proof techniques for proving consistency and normalization.  We provide an
  overview of this prior work (Section~\ref{sec:relatedwork}) and also give an
  in-depth explanation on how various design decisions affect the size of our
  proof and its extensibility to additional features
  (Section~\ref{sec:discuss}).
\end{itemize}

The result of our work is an artifact that an interested researcher can
navigate and understand. We accompany this short mechanized proof with an
informal description, presented here using set-theoretic notation and
terminology so that it is accessible to readers with a general mathematical
background.  That said, our explanations do not stray too far away from our
proof scripts.  We link each lemma directly to its counterpart in the proof
script, anticipating that readers may wish to see how these results may be
expressed and verified in a proof assistant.  The typeset proofs purposefully
follow the mechanized proofs closely while avoiding, as much as possible,
artifacts specific to Coq.

Not only does this close connection aid readers that wish to, like us, adopt
proof assistants for their day-to-day use, but we also find that this
precision is important for conveying the proof technique itself. Unlike
properties that are derivable through syntactic means, proofs by logical
relations make demands on the strength of the metalogic in which they are
expressed. An informal proof that attempts to be agnostic or ambiguous about
the underlying metatheory requires substantial effort from the reader to
understand whether it is definable in a given ambient logic.

@@

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{refs}

%%
%% If your work has an appendix, this is the place to put it.

\end{document}
\endinput
%%
%% End of file `sample-acmsmall.tex'.

>>>>

\documentclass[\ifpublic nolinenum\else\fi,online,OA]{jfp}

\let\Bbbk\relax % amsmath tries to redefine this
\usepackage{amsthm,amssymb}
\let\draftnote\relax % this is already defined in JFP
\usepackage{draft}
\usepackage[outputdir=latex.out]{minted} % latexrun puts everything in latex.out
\usepackage{xspace,stmaryrd,bbding}
\usepackage[para]{footmisc}
\usepackage{ottalt}
\inputott{rules}

\ifcomments
\newnote{scw}{blue} % Stephanie Weirich
\newnote{yl}{purple} % Yiyun Liu
\newnote{jc}{darkgreen}
\else
\newcommand{\scw}[1]{}
\newcommand{\yl}[1]{}
\newcommand{\jc}[1]{}
\fi

\newcommand{\dotv}[2]{\href{#1}{\texttt{#1}}{\texttt{:#2}}}
\newcommand{\lang}{$\lambda^{\Pi}$\xspace}

@@

\citestyle{acmauthoryear}

\begin{document}

\journaltitle{JFP}
\volno{35}
\edno{0}
\cpr{Cambridge University Press}
\doival{10.1017/}
\totalpg{\pageref{lastpage}}
\jnlDoiYr{2025}
\lefttitle{Y. Liu et al.}
\righttitle{Logical Relations for Type Theory}

\title{Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories}

\begin{authgrp}
\author{Yiyun Liu}
% \orcid{0009-0006-8717-2498}
\affiliation{University of Pennsylvania (\email{\elink{liuyiyun@seas.upenn.edu}})}

% \author{Jonathan Chan}
% \orcid{0000-0003-0830-3180}
% \affiliation{University of Pennsylvania (\email{\elink{jcxz@seas.upenn.edu}})}

\author{Stephanie Weirich}
% \orcid{0000-0002-6756-9168}
\affiliation{University of Pennsylvania (\email{\elink{sweirich@seas.upenn.edu}})}
\end{authgrp}

\begin{abstract}
Proof by logical relations is a powerful technique that has been used
to derive metatheoretic properties of type systems, such as
consistency and parametricity. While there exists a
plethora of introductory materials about logical relations in the
context of simply typed or polymorphic lambda calculi, a streamlined
presentation of proof by logical relation for a dependently typed language
is lacking. In this paper, we present a short
consistency proof for a dependently typed language that contains a
rich set of features, including a countable universe
hierarchy, booleans, and an intensional identity type. We show that
the logical relation can be easily extended to prove the existence of
$\beta\eta$-normal forms.
We have
fully mechanized the consistency proof using the Coq proof assistant
in under 1000 lines of code, with 500 lines of additional code for the
$\beta\eta$-normal form extension.
\end{abstract}

\maketitle[T]

\section{Introduction}
This paper presents a \emph{short} and \emph{mechanized} proof of
logical consistency for \lang{}, a dependent type theory with a
predicative universe hierarchy, large elimination, an intensional identity
type, a boolean base type, and dependent elimination forms.

@@

However, tutorial material on syntactic logical
relations~\citep{skorstengaard2019introduction, harpertait, harperkripke,
  pierce2002types, pierce2004advanced,harper2016practical} is primarily
focused on simple or polymorphic types. In that context,
syntactic logical relations can be defined as simple recursive functions over
the structure of types, or in the case of recursive types, defined over the
evaluation steps of the computation. Yet neither of these techniques can be
used to define a logical relation in the context of a predicative dependent
type theory, so a novice researcher may have the impression
that logical relations are not applicable to dependent types.

But this is not the case. 
% \scw{Add a sentence about pen-and-paper LR proofs
%  for dependent type theories?}\yl{There is a long paragraph in Sec 8 about how
%  pnp proofs like Geuvers' can't be easily mechanized, but I'm not sure if we
%  can also argue that they are hard to read, too (I personally think that's the
%  case, but I don't know if the reviewers will agree)} 
Recent authors have developed tour-de-force
mechanizations for the metatheory of modern proof
assistants~\citep{nbeincoq,decagda,martin-lof-a-la-coq,anand2014towards}, and
rely on logical relations as part of their
developments. However, because these proofs show diverse results about real
systems and algorithms, these developments range in size from 10,000 to
300,000 lines of code. As a result, their uses of logical relations are
difficult to isolate from the surrounding contexts and inaccessible
to casual readers.

Our paper instead provides a gentle and accessible introduction to a powerful
technique for dependent type theories.
It is accompanied by a short mechanized proof development
of fewer than 1000 lines of code, developed using the Coq proof
assistant~\citep{coq}.
We have streamlined our proof through a number of means: the careful selection
of the features that we include in the object type system and the results that
we prove about it, in addition to the judicious use of automation.

Our language is small, but includes enough to be illustrative. For example, we
eschew inductive datatypes and W types, but we do include propositional
equality and booleans to capture the challenges presented by indexed types and
dependent pattern matching. We do not show the decidability of type checking,
nor do we develop a PER semantics, but we prove logical consistency, which
states that empty types are not inhabited in an empty context,
and extend our consistency proof to show the existence of $\beta\eta$-normal
forms for well-typed closed \emph{and} open terms,
at a moderate cost of 600 lines of code.
We include a full predicative universe hierarchy with large elimination
to demonstrate the logical strength of the approach.
% JC: I changed "type-level computation" to "large elimination"
% since I think it's usually the latter that gets referred to
% when talking about logical strength and expressivity.

@@

Concretely, our paper makes the following contributions.
\begin{itemize}
\item In Section~\ref{sec:spec}, we introduce \lang{}, the dependent type
  theory of interest. A key design choice that impacts our proofs is the use
  of an untyped conversion rule, inspired by Pure Type
  Systems~\citep{barendregt1991introduction}, and specified through parallel
  reduction~\citep{takahashi-parallel-reduction,
    barendregt:lambda-calculi-with-types}.
\item In Section~\ref{sec:logreldep}, we formulate logical consistency for
  \lang{}, the property of interest, to motivate a logical relation. We
  define the relation inductively, prove its functionality,
  define semantic typing in terms of the relation, and prove the
  fundamental theorem, from which consistency follows as a corollary
  (Section~\ref{sec:logrelproof}). Our proof
  showcases the special treatment required to model many common
  features of dependent type theories, thus making our proof applicable to a
  broad range of type systems.
\item We strengthen our logical relation to prove the existence of
  $\beta$-normal forms (Section~\ref{sec:extension}) and $\beta\eta$-normal
  forms (Section~\ref{sec:betaeta}) for well-typed open terms. The
  modifications made to our initial logical relation are small and closely
  mirror the necessary extensions for a simply typed language. This shows
  that once we have established the base techniques, we can port
  ideas from proofs about simpler languages to the dependently typed setting.
\item We mechanize all our proofs in Coq, with 983
  lines of code for the consistency proof with $\beta$ rules and a moderate
  increase to 1596 lines of code for the normalization proof with $\beta\eta$
  rules. We discuss our choice of using Coq as our metatheory,
  including the use of off-the-shelf
  semantic engineering infrastructure and automation tools, in
  Section~\ref{sec:logrelmech}. Our proof development is available as
  supplementary material.
\item We compare our work to existing proofs by logical relations and other
  proof techniques for proving consistency and normalization.  We provide an
  overview of prior work (Section~\ref{sec:relatedwork}) and explain
  how various design decisions affect the size of our
  proof and its extensibility to additional features
  (Section~\ref{sec:discuss}).
\end{itemize}

The result of our work is an artifact targeted towards researchers
familiar with the syntax of dependent types, as well as logical relations
for simple or polymorphic types, and who wish to understand using logical
relations \emph{for} dependent types.
The short mechanized proof is accompanied here by a pen-and-paper description
using set-theoretic notation and terminology so that it is accessible to
readers with a general mathematical background. This pen-and-paper proof
purposefully follows the mechanized proof closely while avoiding Coq-specific
details as much as possible, and lemmas are linked directly to their
counterparts in the mechanization.

Not only does this close connection aid readers who wish to adopt proof
assistants for mechanizing metatheory, this precision is also important for
conveying the proof technique itself. Unlike properties derivable through
purely syntactic means, proofs by logical relations make demands on the
strength of the metatheory in which they are expressed. An informal proof that
attempts to be agnostic or ambiguous about the underlying metatheory requires
substantial effort from the reader to understand whether it is definable in a
given ambient logic.

@@

\bibliographystyle{ACM-Reference-Format}
\bibliography{refs}

\label{lastpage}

\end{document}
